Query Optimization for Dynamic Imputation

### 2017-05-15 paper/relatedwork.tex
paper: stop using citations as nouns

Imputation of missing values is widely studied in the statistics and machine
learning communities. As highlighted in~\cite{gelman2006data}, missing data
can appear for a variety of reasons.
It can be missing uniformly at random or conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between attributes to account for different forms
of missingness. For example, \cite{burgette2010multiple} discuss the
usage of iterative decision trees for imputing missing data.

\cite{akande2015empirical} analyzes the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well-known and can limit approaches. For example, \cite{akande2015empirical} finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a trade-off between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
provided a treatment of the semantics of \nullv{}. Multiple
papers have described various (at times conflicting) treatments
of \nullv{}s~\cite{grant1977null}. \ProjectName's main design invariant---no relational operator
sees missing values for attributes it must operate on nor do users see
missing data---eliminates
the need to handle \nullv{} value semantics, while guaranteeing query evaluation soundness.

Database system developers and others have worked on techniques to automatically
detect dirty values, whether missing or otherwise, and rectify the errors if
possible. A survey of methods and systems is provided in~\cite{hellerstein2008quantitative}.

In~\cite{wolf2007query}, queries over databases with missing values are
processed directly using a statistical approach, taking advantage of correlations between
attributes. The tuples that match the original query as well as a ranking of
tuples with incomplete data that may match are returned. Our work differs in
that we allow any well-formed statistical technique to be used for imputation
and focus on returning results to the analyst as if the database had been
complete.

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. In~\cite{fernandez2009inter}, feedback punctuation is used to dynamically
reconfigure the query plan for state-dependent
optimizations as data continues to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.

@@

\review{
\cite{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
of probabilistic compositional repair operators, called Lenses. These can be used
to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
performed (expressed by the confidence of their output) and the cost of those operations. However, their 
repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
variables with inferred values. Meanwhile, \ProjectName{} optimizes imputation at the plan-level and
provides the same impute technique for all tuples streaming through a given operator.
}

@@

\review{
The planning algorithm that we present has similarities to the multi-objective query optimizer in~\cite{trummer2014}.
Both algorithms handle plans which have multiple cost functions by maintaining sets of Pareto optimal plans and pruning plans which become dominated.
In \ProjectName{}, costs are expected to grow monotonically as the size of the plan increases, which simplifies the optimization problem significantly.
\cite{trummer2014} handles the more complex case of cost functions which are piecewise linear, but not necessarily monotonic.
}

\subsection{Forecasting and Databases}
\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In their follow on work~\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

\cite{fischer2013towards} describe the architecture of a DBMS with integrated forecasting operations for time-series, detailing
the abstractions necessary to do so.

In contrast to this work, \ProjectName{} is targeted at generic value imputation, not necessarily tailored to 
time-series. The optimizer is not based on equivalence transformations, nor are there guarantees of equal
results under different conditions. Instead, the optimizer allows users to pick their trade-off between
runtime cost and imputation quality. The search space considered by our optimizer is broader, not just
forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan
(with some restrictions). The novelty of our contribution lies in the successful incorporation of
imputation operations in non-trivial query plans with cost-based optimization.

\cite{duan2007processing} describes the Fa system and its declarative language for time-series forecasting. Their
system automatically searches the space of attribute combinations/transformations and statistical models
to produce forecasts within a given accuracy threshold. Accuracy estimates are determined using
standard techniques, such as cross-validation.

>>>>

Imputation of missing values is widely studied in the statistics and machine
learning communities. Missing data
can appear for a variety of reasons~\cite{gelman2006data}.
It can be missing uniformly at random or conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between attributes to account for different forms
of missingness.
For example, \textcite*{burgette2010multiple} use iterative decision trees for imputing missing data.

The computational difficulties of imputing on large base
tables are well-known and can limit approaches. % \citeauthor{akande2015empirical} analyze the performance of various
% multiple imputation techniques on the American Community Survey dataset~\cite{akande2015empirical}.
For example, \textcite*{akande2015empirical} find that one approach (MI-GLM) is prohibitively expensive when attempting to impute on the American Community Survey dataset.
In this case, the complexity comes from the large domains of the variables in the ACS data (up to ten categories in their case).
In contrast, \ProjectName{} allows users to specify a trade-off between
imputation quality and runtime performance, allowing users to perform queries
directly on datasets which are too large to fully impute quickly. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of \nullv{}.
Multiple papers have described various (at times conflicting) treatments of \nullv{}s~\cite{codd1973understanding,grant1977null}.
\ProjectName's main design invariant---no relational operator
sees missing values for attributes it must operate on nor do users see
missing data---eliminates
the need to handle \nullv{} value semantics, while guaranteeing query evaluation soundness.

Database system developers and others have worked on techniques to automatically
detect dirty values, whether missing or otherwise, and rectify the errors if
possible. \textcite*{hellerstein2008quantitative} surveys the methods and systems related to missing value detection and correction.

\textcite*{wolf2007query} treat queries over databases with missing values directly using a statistical approach, taking advantage of correlations between
attributes. The tuples that match the original query as well as a ranking of
tuples with incomplete data that may match are returned. Our work differs in
that we allow any well-formed statistical technique to be used for imputation
and focus on returning results to the analyst as if the database had been
complete.

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. \todo{What does this mean?} \textcite*{fernandez2009inter} use feedback punctuation to dynamically
reconfigure the query plan for state-dependent
optimizations as data continues to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.

@@

\review{
\textcite*{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
of probabilistic compositional repair operators, called Lenses. These can be used
to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
performed (expressed by the confidence of their output) and the cost of those operations. However, their 
repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
variables with inferred values. In contrast, \ProjectName{} optimizes imputation at the plan-level and
provides the same impute technique for all tuples streaming through a given operator.
}

@@

\review{
The planning algorithm that we present has similarities to work on multi-objective query optimization~\cite{trummer2014}.
Both algorithms handle plans which have multiple cost functions by maintaining sets of Pareto optimal plans and pruning plans which become dominated.
In \ProjectName{}, costs are expected to grow monotonically as the size of the plan increases, which simplifies the optimization problem significantly.
\textcite*{trummer2014} handle the more complex case of cost functions which are piecewise linear, but not necessarily monotonic.
}

\subsection{Forecasting and Databases}
\textcite*{parisi2011embedding} introduces the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In subsequent work~\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

\textcite*{fischer2013towards} describe the architecture of a DBMS with integrated forecasting operations for time-series, detailing
the abstractions necessary to do so.

In contrast to this work, \ProjectName{} is targeted at generic value imputation, and is not tailored to time-series.
The optimizer is not based on equivalence transformations, nor are there guarantees of equal
results under different conditions. Instead, the optimizer allows users to pick their trade-off between
runtime cost and imputation quality. The search space considered by our optimizer is broader, not just
forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan
(with some restrictions). The novelty of our contribution lies in the successful incorporation of
imputation operations in non-trivial query plans with cost-based optimization.

\textcite*{duan2007processing} describe the Fa system and its declarative language for time-series forecasting. Their
system automatically searches the space of attribute combinations/transformations and statistical models
to produce forecasts within a given accuracy threshold. Accuracy estimates are determined using
standard techniques, such as cross-validation.
